---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cheapr

<!-- badges: start -->

[![R-CMD-check](https://github.com/NicChr/cheapr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NicChr/cheapr/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/cheapr)](https://CRAN.R-project.org/package=cheapr)
<!-- badges: end -->

In cheapr, 'cheap' means fast and memory-efficient, and that's exactly the philosophy that cheapr aims to follow.

## Installation

You can install the development version of cheapr like so:

``` r
remotes::install_github("NicChr/cheapr")
```

## Last-observation carried forward (minor optimisation)

`num_na()` is a useful function to efficiently return the number of `NA` values and can be used in a variety of problems.

Here is an example of a minor optimisation we can add to `vctrs::vec_fill_missing` to return x if x has zero or only `NA` values.

```{r example}
library(cheapr)
library(vctrs)
library(bench)

na_locf <- function(x){
  # num_na is recursive so we compare it to unlisted length
  if (num_na(x) %in% c(0, unlisted_length(x))){
    x
  } else {
    vec_fill_missing(x, direction = "down")
  }
}
x <- rep(NA, 10^6)
identical(x, na_locf(x))
mark(na_locf(x), vec_fill_missing(x, direction = "down"))
mark(na_locf(x), vec_fill_missing(x, direction = "down"))
```

All the `NA` handling functions in cheapr can make use of multiple cores on your machine using openMP.

```{r}
# 1 core by default
mark(num_na(x), sum(is.na(x)))
# 4 cores
options(cheapr.cores = 4)
mark(num_na(x), sum(is.na(x)))
```

## Efficient NA counts by row/col

```{r}
m <- matrix(x, ncol = 10^3)
# Number of NA values by row
mark(row_na_counts(m), 
     rowSums(is.na(m)))
# Number of NA values by col
mark(col_na_counts(m), 
     colSums(is.na(m)))
```

`is_na` is a multi-threaded alternative to `is.na` 

```{r}
x <- rnorm(10^6)
x[sample.int(10^6, 10^5)] <- NA
mark(is.na(x), is_na(x))

### posixlt method is much faster
hours <- as.POSIXlt(seq.int(0, length.out = 10^6, by = 3600),
                    tz = "UTC")
hours[sample.int(10^6, 10^5)] <- NA

mark(is.na(hours), is_na(hours))
```

It differs in 2 regards:

* List elements are regarded as `NA` when either that element is an `NA` value 
or it is a list containing only `NA` values.
* For data frames, `is_na` returns a logical vector where `TRUE` defines an empty 
row of only `NA` values.

```{r}
# List example
is.na(list(NA, list(NA, NA), 10))
is_na(list(NA, list(NA, NA), 10))

# Data frame example
df <- data.frame(x = c(1, NA, 3),
                 y = c(NA, NA, NA))
df
is_na(df)
is_na(df)
# The below identity should hold
identical(is_na(df), row_na_counts(df) == ncol(df))
```

`is_na` and all the `NA` handling functions fall back on calling `is.na()` if no 
suitable method is found. This means that custom objects like vctrs rcrds and
more are supported.

## Cheap data frame summaries with `overview`

Inspired by the excellent skimr package, `overview()` is a cheaper alternative
designed for larger data.

```{r}
df <- data.frame(
  x = sample.int(100, 10^7, TRUE),
  y = factor_(sample(LETTERS, 10^7, TRUE)),
  z = rnorm(10^7)
)
overview(df, hist = TRUE)
mark(overview(df))
```


## Greatest common divisor and smallest common multiple

```{r}
gcd2(5, 25)
scm2(5, 6)

gcd(seq(5, 25, by = 5))
scm(seq(5, 25, by = 5))

x <- seq(1L, 1000000L, 1L)
mark(gcd(x))
x <- seq(0, 10^6, 0.5)
mark(gcd(x))
```

## Creating many sequences

As an example, to create 3 sequences with different increments,\
the usual approach might be to use lapply to loop through the increment values together with `seq()`

```{r}
# Base R
increments <- c(1, 0.5, 0.1)
start <- 1
end <- 5
unlist(lapply(increments, \(x) seq(start, end, x)))
  
```

In cheapr you can use `seq_()` which accepts vector arguments.

```{r}
seq_(start, end, increments)
```

Use `add_id = TRUE` to label the individual sequences.

```{r}
seq_(start, end, increments, add_id = TRUE)
```

If you know the sizes of your sequences beforehand, use `sequence_()`

```{r}
seq_sizes <- c(3, 5, 10)
sequence_(seq_sizes, from = 0, by = 1/3, add_id = TRUE) |> 
  enframe_()
```

You can also calculate the sequence sizes using `seq_size()`

```{r}
seq_size(start, end, increments)
```

## 'Cheaper' Base R alternatives

### which

```{r}
# which()
x <- rep(TRUE, 10^6)
mark(cheapr_which = which_(x),
     base_which = which(x))
x <- rep(FALSE, 10^6)
mark(cheapr_which = which_(x),
     base_which = which(x))
x <- c(rep(TRUE, 5e05), rep(FALSE, 1e06))
mark(cheapr_which = which_(x),
     base_which = which(x))
x <- c(rep(FALSE, 5e05), rep(TRUE, 1e06))
mark(cheapr_which = which_(x),
     base_which = which(x))
x <- sample(c(TRUE, FALSE), 10^6, TRUE)
x[sample.int(10^6, 10^4)] <- NA
mark(cheapr_which = which_(x),
     base_which = which(x))
```

### factor

```{r}
# factor()
x <- sample(seq(-10^3, 10^3, 0.01))
y <- do.call(paste0, expand.grid(letters, letters, letters, letters))
mark(cheapr_factor = factor_(x), 
     base_factor = factor(x))
mark(base_factor = factor_(x, order = FALSE), 
     base_factor = factor(x, levels = unique(x)))
mark(cheapr_factor = factor_(y), 
     base_factor = factor(y))
mark(base_factor = factor_(y, order = FALSE), 
     base_factor = factor(y, levels = unique(y)))
```

### intersect & setdiff

```{r}
# intersect() & setdiff()
x <- sample.int(10^6, 10^5, TRUE)
y <- sample.int(10^6, 10^5, TRUE)
mark(cheapr_intersect = intersect_(x, y, dups = FALSE),
     base_intersect = intersect(x, y))
mark(cheapr_setdiff = setdiff_(x, y, dups = FALSE),
     base_setdiff = setdiff(x, y))
```

### `%in_%` and `%!in_%`

```{r}
mark(cheapr = x %in_% y,
     base = x %in% y)
mark(cheapr = x %!in_% y,
     base = !x %in% y)
```

### cut.default

```{r}
# cut.default()
x <- rnorm(10^7)
b <- seq(0, max(x), 0.2)
mark(cheapr_cut = cut_numeric(x, b), 
     base_cut = cut(x, b))
```
